Internal models of type theory

### 2013-11-08 article-models-tt.tex
start leaving infarray environment; compatibility of typeconstructors with idtypes

\subsection{The basic judgments}
The judgments we wish to make are not the standad six judgments for ``$\Gamma$ is a context'', ``$A$ is
a type in context $\Gamma$'' and ``$x$ is a term of $A$ in context $\Gamma$'' and the three equality
judgments coming with them. We have those, but we also add a judgment for ``$i$ is a term of context $\Gamma$''
and ``$i$ and $j$ are equal terms of context $\Gamma$''. Thus, we have eight basic judgments we can make:

\begin{align*}
\jctx*{\Gamma} & \jctxeq*{\Gamma}{\Gamma'}\\
\jtermc*{\Gamma}{i} & \jtermceq*{\Gamma}{i}{j}\\
\jtype*{\Gamma}{A} & \jtypeeq*{\Gamma}{A}{B}\\
\jterm*{\Gamma}{A}{x} & \jtermeq*{\Gamma}{A}{x}{y}.
\end{align*}

@@

\subsection{The basic rules for judgmental equality}
The rules for judgmental equality establish that it is an equivalence relation
in all three cases (contexts, types and terms).
%\begingroup
%\renewcommand*{\arraystretch}{3}
%\begin{equation*}
%\begin{array}{ccc}
\begin{infarray}{ccc}
\inference{\jctx{\Gamma}}{\jctxeq{\Gamma}{\Gamma}} & \inference{\jctxeq{\Gamma}{\Delta}}{\jctxeq{\Delta}{\Gamma}} & \inference{\jctxeq{\Gamma}{\Delta}\qquad\jctxeq{\Delta}{\Theta}}{\jctxeq{\Gamma}{\Theta}}\\
\inference{\jtype{\Gamma}{A}}{\jtypeeq{\Gamma}{A}{A}} &
\inference{\jtypeeq{\Gamma}{A}{B}}{\jtypeeq{\Gamma}{B}{A}} & 
\inference{\jtypeeq{\Gamma}{A}{B}\qquad\jtypeeq{\Gamma}{B}{C}}{\jtypeeq{\Gamma}{A}{C}}\\
\inference{\jterm{\Gamma}{A}{x}}{\jtermeq{\Gamma}{A}{x}{x}} & 
\inference{\jtermeq{\Gamma}{A}{x}{y}}{\jtermeq{\Gamma}{A}{y}{x}} &
\inference{\jtermeq{\Gamma}{A}{x}{y}\qquad\jtermeq{\Gamma}{A}{y}{z}}{\jtermeq{\Gamma}{A}{x}{z}}
\end{infarray}
%\end{array}
%\end{equation*}
\begin{infarray}{cc}
\inference{\jctxeq{\Gamma}{\Delta}\qquad\jtype{\Gamma}{A}}{\jtype{\Delta}{A}}
& \inference{\jctxeq{\Gamma}{\Delta}\qquad\jtypeeq{\Gamma}{A}{B}}{\jtypeeq{\Delta}{A}{B}}\\
\inference{\jctxeq{\Gamma}{\Delta}\qquad\jterm{\Gamma}{A}{x}}{\jterm{\Delta}{A}{x}}
& \inference{\jctxeq{\Gamma}{\Delta}\qquad\jtermeq{\Gamma}{A}{x}{y}}{\jtermeq{\Delta}{A}{x}{y}}\\
\inference{\jtypeeq{\Gamma}{A}{B}\qquad \jterm{\Gamma}{A}{x}}{\jterm{\Gamma}{B}{x}}
& \inference{\jtypeeq{\Gamma}{A}{B}\qquad\jtermeq{\Gamma}{A}{x}{y}}{\jtermeq{\Gamma}{B}{x}{y}}
\end{infarray}

@@

\subsubsection{Substitution is compatible with extension}
Suppose $\jterm{\Gamma}{A}{x}$ in all of the following inference rules.
\begin{infarray}{c}
\inference{\jtype{\ctxext{{\Gamma}{A}}{P}}{Q}}{\jtypeeq{\Gamma}{\subst{x}{\ctxext{P}{Q}}}{\ctxext{\subst{x}{P}}{\subst{x}{Q}}}}
\end{infarray}

@@

\subsection{The variable rule -- identity functions}
The variable rule for contexts:
\begin{infarray}{c}
\inference{\jctx{\Gamma}}{\jterm{\Gamma}{\ctxwk{\Gamma}{\Gamma}}{\idfunc[\Gamma]}}\\
\inference{\jterm{\Gamma}{A}{x}}{\jtermeq{\Gamma}{A}{\subst{\idfunc[\Gamma]}{\ctxwk{\Gamma}{x}}}{x}}\\
\inference{\jterm{\Delta}{\ctxwk{\Delta}{\Gamma}}{g}}{\jtermeq{\Delta}{\ctxwk{\Delta}{\Gamma}}{\subst{g}{\ctxwk{\Delta}{\idfunc[\Gamma]}}}{g}}
\end{infarray}

The variable rule for types:
\begin{infarray}{c}
\inference{\jtype{\Gamma}{A}}{\jterm{\ctxext{\Gamma}{A}}{\ctxwk{A}{A}}{\idfunc[A]}}\\
\inference{\jterm{\Gamma}{A}{x}}{\jtermeq{\Gamma}{A}{\subst{x}{\idfunc[A]}}{x}}\\
\inference{\jtype{\ctxext{\Gamma}{A}}{P}}{\jtypeeq{\ctxext{\Gamma}{A}}{\subst{\idfunc[A]}{\ctxwk{A}{P}}}{P}}\\
\inference{\jterm{\ctxext{\Gamma}{A}}{P}{f}}{\jtermeq{\ctxext{\Gamma}{A}}{P}{\subst{\idfunc[A]}{\ctxwk{A}{f}}}{f}}\\
\inference{\jterm{\ctxext{\Gamma}{B}}{\ctxwk{B}{A}}{g}}{\jtermeq{\ctxext{\Gamma}{B}}{\ctxwk{B}{A}}{\subst{g}{\ctxwk{B}{\idfunc[A]}}}{g}}
\end{infarray}

@@

\subsection{Identity types}
The identity types may now be formulated just as in ordinary Martin\nobreakdash-L\"of
type theory. However, since all model categories have path objects, we will also
introduce identity types for contexts.

@@

\begin{infarray}{c}
\inference{\jctx{\Gamma}}{\jtype{\ctxext{\Gamma}{\ctxwk{\Gamma}{\Gamma}}}{\idtypevar{\Gamma}}}\\
\inference{\jctx{\Gamma}}{\jterm{\Gamma}{\subst{\idfunc[\Gamma]}{\idtypevar{\Gamma}}}{\refl{\Gamma}}}\\
\inference{\jtype{\ctxext{{\Gamma}{\ctxwk{\Gamma}{\Gamma}}}{\idtypevar{\Gamma}}}{P}
           \qquad
           \jterm{\Gamma}{\subst{\refl{\Gamma}}{{\idfunc[\Gamma]}{P}}}{d}}
           {\jterm{\ctxext{{\Gamma}{\ctxwk{\Gamma}{\Gamma}}}{\idtypevar{\Gamma}}}{P}{\tfJ(d)}}\\
\inference{\jtype{\ctxext{{\Gamma}{\ctxwk{\Gamma}{\Gamma}}}{\idtypevar{\Gamma}}}{P}\qquad
\jterm{\Gamma}{\subst{\refl{\Gamma}}{{\idfunc[\Gamma]}{P}}}{d}}
{\jtermeq
  {\ctxext{{\Gamma}{\ctxwk{\Gamma}{\Gamma}}}{\idtypevar{\Gamma}}}
  {\subst{\refl{\Gamma}}{{\idfunc[\Gamma]}{P}}}
  {\subst{\refl{\Gamma}}{{\idfunc[\Gamma]}{\tfJ(d)}}}
  {d}}\\
\inference{\jtype{\ctxext{{{\Gamma}{\ctxwk{\Gamma}{\Gamma}}}{\idtypevar{\Gamma}}}{P}}{Q}
           \qquad
           \jterm{\ctxext{\Gamma}{\subst{\refl{\Gamma}}{{\idfunc[\Gamma]}{P}}}}{\subst{\refl{\Gamma}}{{\idfunc[\Gamma]}{Q}}}{d}}
           {\jterm{\ctxext{{{\Gamma}{\ctxwk{\Gamma}{\Gamma}}}{\idtypevar{\Gamma}}}{P}}{Q}{\tfJ(d)}}\\
\inference
  {\jtype
    {\ctxext{{{\Gamma}{\ctxwk{\Gamma}{\Gamma}}}{\idtypevar{\Gamma}}}{P}}
    {Q}
  \qquad
  \jterm
    {\ctxext{\Gamma}{\subst{\refl{\Gamma}}{{\idfunc[\Gamma]}{P}}}}
    {\subst{\refl{\Gamma}}{{\idfunc[\Gamma]}{Q}}}
    {d}}
  {\jtermeq
    {\ctxext{{{\Gamma}{\ctxwk{\Gamma}{\Gamma}}}{\idtypevar{\Gamma}}}{P}}
    {\subst{\refl{\Gamma}}{{\idfunc[\Gamma]}{Q}}}
    {\subst{\refl{\Gamma}}{{\idfunc[\Gamma]}{\tfJ(d)}}}
    {d}}
\end{infarray}

@@

\begin{infarray}{c}
\inference{\jtype{\Gamma}{A}}{\jtype{\ctxext{{\Gamma}{A}}{\ctxwk{A}{A}}}{\idtypevar{A}}}\\
\inference{\jtype{\Gamma}{A}}{\jterm{\ctxext{\Gamma}{A}}{\subst{\idfunc[A]}{\idtypevar{A}}}{\refl{A}}}\\
\inference{\jtype{\ctxext{{{\Gamma}{A}}{\ctxwk{A}{A}}}{\idtypevar{A}}}{P}\qquad
\jterm{\ctxext{\Gamma}{A}}{\subst{\refl{A}}{{\idfunc[A]}{P}}}{d}}
{\jterm{\ctxext{{{\Gamma}{A}}{\ctxwk{A}{A}}}{\idtypevar{A}}}{P}{\tfJ(d)}}\\
\inference{\jtype{\ctxext{{{\Gamma}{A}}{\ctxwk{A}{A}}}{\idtypevar{A}}}{P}\qquad
\jterm{\ctxext{\Gamma}{A}}{\subst{\refl{A}}{{\idfunc[A]}{P}}}{d}}
{\jtermeq
  {\ctxext{{{\Gamma}{A}}{\ctxwk{A}{A}}}{\idtypevar{A}}}
  {\subst{\refl{A}}{{\idfunc[A]}{P}}}
  {\subst{\refl{A}}{{\idfunc[A]}{\tfJ(d)}}}
  {d}}
\end{infarray}

@@

@@

\begin{infarray}{c}
\inference{}{\jctx{\unit}}\\
\inference{\jctx{\Gamma}}{\jhom{\Gamma}{\unit}{\tounit{\Gamma}}}\\
\inference{\jctx{\Gamma}\qquad\jhom{\Gamma}{\unit}{f}}{\jhomeq{\Gamma}{\unit}{f}{\tounit{\Gamma}}}
\end{infarray}

@@

\begin{infarray}{c}
\inference{\jctx{\Gamma}}{\jtype{\Gamma}{\unit_\Gamma}}\\
\inference{\jctx{\Gamma}}{\jterm{\Gamma}{\unit_\Gamma}{\ttt_\Gamma}}\\
\inference{\jtype{\ctxext{\Gamma}{\unit_\Gamma}}{P}\qquad\jterm{\Gamma}{\subst{\ttt_\Gamma}{P}}{u}}
          {\jterm{\ctxext{\Gamma}{\unit_\Gamma}}{P}{\ind{\unit_\Gamma}(u)}}\\
\inference{\jtype{\ctxext{\Gamma}{\unit_\Gamma}}{P}\qquad\jterm{\Gamma}{\subst{\ttt_\Gamma}{P}}{u}}
          {\jtermeq{\Gamma}{\subst{\ttt_\Gamma}{P}}{\subst{\ttt_\Gamma}{\ind{\unit_\Gamma}(u)}}{u}}
\end{infarray}

@@

\begin{defn}
\begin{infarray}{c}
\inference{\jctx{\Gamma}}{\jctx{\tau\Gamma}}\\
\inference{\jctx{\Gamma}}{\jhom{\Gamma}{\tau\Gamma}{t}}\\
\inference{\jctx{\Gamma}}{\jtermeq{\ctxext{\Gamma}{\ctxwk{\Gamma}{\Gamma}}}{\ctxwk{{\Gamma}{\Gamma}}{{\Gamma}{\Delta}}}{\ctxwk{\Gamma}{t}}{\ctxwk{{\Gamma}{\Gamma}}{t}}}\\
\inference{\jhom{\Gamma}{\Delta}{f}\qquad\jtermeq{\ctxext{\Gamma}{\ctxwk{\Gamma}{\Gamma}}}{\ctxwk{{\Gamma}{\Gamma}}{{\Gamma}{\Delta}}}{\ctxwk{\Gamma}{f}}{\ctxwk{{\Gamma}{\Gamma}}{f}}}{\jhom{\tau\Gamma}{\Delta}{\tilde{f}}}\\
\inference{\jhom{\Gamma}{\Delta}{f}\qquad\jtermeq{\ctxext{\Gamma}{\ctxwk{\Gamma}{\Gamma}}}{\ctxwk{{\Gamma}{\Gamma}}{{\Gamma}{\Delta}}}{\ctxwk{\Gamma}{f}}{\ctxwk{{\Gamma}{\Gamma}}{f}}}
{\jtermeq{\Gamma}{\ctxwk{\Gamma}{\Delta}}{\jcomp{\Gamma}{t}{\tilde{f}}}{f}}
\end{infarray}
\end{defn}

@@

\subsubsection{Products}
\begin{infarray}{c}
\inference{\jctx{\Gamma}\qquad\jctx{\Delta}}{\jctx{\product{\Gamma}{\Delta}}}\\
\inference{\jctx{\Gamma}\qquad\jctx{\Delta}}{\jhom{{\Gamma}{\Delta}}{\product{\Gamma}{\Delta}}{\pair}}
\end{infarray}

@@

\begin{infarray}{cc}
\inference{\jctx{\Gamma}\qquad\jctx{\Delta}}{\jctx{\product{\Gamma}{\Delta}}}
 & \inference{\jhom{\greek{E}}{\Gamma}{f}\qquad\jhom{\greek{E}}{\Delta}{g}}
          {\jhom{\greek{E}}{\product{\Gamma}{\Delta}}{\pairp{f,g}}}\\
\inference{\jctx{\Gamma}\qquad\jctx{\Delta}}{\jhom{\product{\Gamma}{\Delta}}{\Gamma}{\proj1}}
 & \inference{\jhom{\greek{E}}{\Gamma}{f}\qquad\jhom{\greek{E}}{\Delta}{g}}
          {\jhomeq{\greek{E}}{\Gamma}{\jcomp{\greek{E}}{\pairp{f,g}}{\proj1}}{f}}\\
\inference{\jctx{\Gamma}\qquad\jctx{\Delta}}{\jhom{\product{\Gamma}{\Delta}}{\Delta}{\proj2}} 
 & \inference{\jhom{\greek{E}}{\Gamma}{f}\qquad\jhom{\greek{E}}{\Delta}{g}}
          {\jhomeq{\greek{E}}{\Delta}{\jcomp{\greek{E}}{\pairp{f,g}}{\proj2}}{g}}
\end{infarray}
\begin{equation*}
\inference{\jhom{\greek{E}}{\product{\Gamma}{\Delta}}{h}
           \qquad
           \jhomeq{\greek{E}}{\Gamma}{\jcomp{E}{h}{\proj1}}{f}
           \qquad
           \jhomeq{\greek{E}}{\Delta}{\jcomp{E}{h}{\proj2}}{g}}
          {\jhomeq{\greek{E}}{\product{\Gamma}{\Delta}}{h}{\pairp{f,g}}}
\end{equation*}

@@

>>>>

Perhaps a bigger gain is that we obtain an algebraic theory for categories with
fibrations and sections.

\subsection{The basic judgments}
The judgments we wish to make are the standad six judgments for ``$\Gamma$ is a context'', ``$A$ is
a type in context $\Gamma$'' and ``$x$ is a term of $A$ in context $\Gamma$'' and the three equality
judgments coming with them.

\begin{align*}
\jctx*{\Gamma} & \jctxeq*{\Gamma}{\Gamma'}\\
\jtype*{\Gamma}{A} & \jtypeeq*{\Gamma}{A}{B}\\
\jterm*{\Gamma}{A}{x} & \jtermeq*{\Gamma}{A}{x}{y}.
\end{align*}

@@

\subsection{The basic rules for judgmental equality}
The rules for judgmental equality establish that it is an equivalence relation
in all three cases (contexts, types and terms).
\begin{align}
& \inference{\jctx{\Gamma}}{\jctxeq{\Gamma}{\Gamma}} 
& & \inference{\jctxeq{\Gamma}{\Delta}}{\jctxeq{\Delta}{\Gamma}} 
& & \inference{\jctxeq{\Gamma}{\Delta}\qquad\jctxeq{\Delta}{\Theta}}{\jctxeq{\Gamma}{\Theta}}\\
& \inference{\jtype{\Gamma}{A}}{\jtypeeq{\Gamma}{A}{A}} 
& & \inference{\jtypeeq{\Gamma}{A}{B}}{\jtypeeq{\Gamma}{B}{A}}
& & \inference{\jtypeeq{\Gamma}{A}{B}\qquad\jtypeeq{\Gamma}{B}{C}}{\jtypeeq{\Gamma}{A}{C}}\\
& \inference{\jterm{\Gamma}{A}{x}}{\jtermeq{\Gamma}{A}{x}{x}}
& & \inference{\jtermeq{\Gamma}{A}{x}{y}}{\jtermeq{\Gamma}{A}{y}{x}}
& & \inference{\jtermeq{\Gamma}{A}{x}{y}\qquad\jtermeq{\Gamma}{A}{y}{z}}{\jtermeq{\Gamma}{A}{x}{z}}
\end{align}

\begin{align}
& \inference{\jctxeq{\Gamma}{\Delta}\qquad\jtype{\Gamma}{A}}{\jtype{\Delta}{A}}
& & \inference{\jctxeq{\Gamma}{\Delta}\qquad\jtypeeq{\Gamma}{A}{B}}{\jtypeeq{\Delta}{A}{B}}\\
& \inference{\jctxeq{\Gamma}{\Delta}\qquad\jterm{\Gamma}{A}{x}}{\jterm{\Delta}{A}{x}}
& & \inference{\jctxeq{\Gamma}{\Delta}\qquad\jtermeq{\Gamma}{A}{x}{y}}{\jtermeq{\Delta}{A}{x}{y}}\\
& \inference{\jtypeeq{\Gamma}{A}{B}\qquad \jterm{\Gamma}{A}{x}}{\jterm{\Gamma}{B}{x}}
& & \inference{\jtypeeq{\Gamma}{A}{B}\qquad\jtermeq{\Gamma}{A}{x}{y}}{\jtermeq{\Gamma}{B}{x}{y}}
\end{align}

@@

\subsubsection{Substitution is compatible with extension}
Suppose $\jterm{\Gamma}{A}{x}$ in all of the following inference rules.
\begin{infarray}{c}
\inference{\jtype{\ctxext{{\Gamma}{A}}{P}}{Q}}
  {\jtypeeq{\Gamma}{\subst{x}{\ctxext{P}{Q}}}{\ctxext{\subst{x}{P}}{\subst{x}{Q}}}}\\
\inference{\jtype{\ctxext{{{\Gamma}{A}}{P}}{Q}}{R}}
  {\jtypeeq{\ctxext{\Gamma}{\subst{x}{P}}}{\subst{x}{\ctxext{Q}{R}}}{\ctxext{\subst{x}{Q}}{\subst{x}{R}}}}
\end{infarray}

@@

\subsubsection{Extension is compatible with weakening}

\begin{equation*}
\jterm{\ctxext{\Gamma}{\ctxwk{\Gamma}{\ctxext{\Delta}{B}}}}{\ctxwk{{\Gamma}{\ctxext{\Delta}{B}}}{\ctxext{\ctxwk{\Gamma}{\Delta}}{\ctxwk{\Gamma}{B}}}}{ac}
\end{equation*}

\subsection{The identity functions}
Without a rule explicitly asserting the existence of an identity morphism we don't
get one, hence we do that here. The identity morphism is a term which introduced
in ordinary type theory via the variable rule. The variable rule is a bit more
general: it asserts that
\begin{equation*}
\jterm{\Gamma,\,x_1:A_1,\ldots,\,x_n:A_n}{A_i}{x_i}
\end{equation*}
for every $1\leq i\leq n$. Thus, it establishes the projections. In our setting,
we get the projections from the identity morphisms together with weakening. We
already have weakening, so here it suffices to introduce the identity morphisms.
They are determined by their behavior with respect to substitution combined with
weakening.

\begin{align}
&\inference{\jctx{\Gamma}}{\jterm{\Gamma}{\ctxwk{\Gamma}{\Gamma}}{\idfunc[\Gamma]}}\label{ctxIdFuncForm}\\
&\inference{\jtype{\Gamma}{A}}{\jtypeeq{\Gamma}{\subst{\idfunc[\Gamma]}{\ctxwk{\Gamma}{A}}}{A}}\label{ctxIdFuncPbInvType}\\
&\inference{\jterm{\Gamma}{A}{x}}{\jtermeq{\Gamma}{A}{\subst{\idfunc[\Gamma]}{\ctxwk{\Gamma}{x}}}{x}}\label{ctxIdFuncPbInvTerm}\\
&\inference{\jterm{\Delta}{\ctxwk{\Delta}{\Gamma}}{g}}{\jtermeq{\Delta}{\ctxwk{\Delta}{\Gamma}}{\subst{g}{\ctxwk{\Delta}{\idfunc[\Gamma]}}}{g}}\label{ctxIdFuncSubstInvTerm}
\end{align}

The identity morphisms for types are introduced analogously to the identity
morphisms of contexts, the only difference being that we have a rule asserting
that $\subst{x}{\idfunc[A]}\jdeq x$ for every term $x:A$ in context $\Gamma$.

\begin{align}
& \inference{\jtype{\Gamma}{A}}{\jterm{\ctxext{\Gamma}{A}}{\ctxwk{A}{A}}{\idfunc[A]}}\\
& \inference{\jterm{\Gamma}{A}{x}}{\jtermeq{\Gamma}{A}{\subst{x}{\idfunc[A]}}{x}}\\
& \inference{\jtype{\ctxext{\Gamma}{A}}{P}}{\jtypeeq{\ctxext{\Gamma}{A}}{\subst{\idfunc[A]}{\ctxwk{A}{P}}}{P}}\\
& \inference{\jterm{\ctxext{\Gamma}{A}}{P}{f}}{\jtermeq{\ctxext{\Gamma}{A}}{P}{\subst{\idfunc[A]}{\ctxwk{A}{f}}}{f}} \\
& \inference{\jterm{\ctxext{\Gamma}{B}}{\ctxwk{B}{A}}{g}}{\jtermeq{\ctxext{\Gamma}{B}}{\ctxwk{B}{A}}{\subst{g}{\ctxwk{B}{\idfunc[A]}}}{g}}
\end{align}

@@

\section{Identity types}
The identity types may now be formulated just as in ordinary Martin\nobreakdash-L\"of
type theory. However, since all model categories have path objects, we will also
introduce identity types for contexts.

@@

\begin{align}
& \inference{\jctx{\Gamma}}{\jtype{\ctxext{\Gamma}{\ctxwk{\Gamma}{\Gamma}}}{\idtypevar{\Gamma}}}\\
& \inference{\jctx{\Gamma}}{\jterm{\Gamma}{\subst{\idfunc[\Gamma]}{\idtypevar{\Gamma}}}{\refl{\Gamma}}}\\
& \inference{\jtype{\ctxext{{\Gamma}{\ctxwk{\Gamma}{\Gamma}}}{\idtypevar{\Gamma}}}{P}
           \qquad
           \jterm{\Gamma}{\subst{\refl{\Gamma}}{{\idfunc[\Gamma]}{P}}}{d}}
           {\jterm{\ctxext{{\Gamma}{\ctxwk{\Gamma}{\Gamma}}}{\idtypevar{\Gamma}}}{P}{\tfJ(d)}}\\
& \inference{\jtype{\ctxext{{\Gamma}{\ctxwk{\Gamma}{\Gamma}}}{\idtypevar{\Gamma}}}{P}\qquad
\jterm{\Gamma}{\subst{\refl{\Gamma}}{{\idfunc[\Gamma]}{P}}}{d}}
{\jtermeq
  {\ctxext{{\Gamma}{\ctxwk{\Gamma}{\Gamma}}}{\idtypevar{\Gamma}}}
  {\subst{\refl{\Gamma}}{{\idfunc[\Gamma]}{P}}}
  {\subst{\refl{\Gamma}}{{\idfunc[\Gamma]}{\tfJ(d)}}}
  {d}}\\
& \inference{\jtype{\ctxext{{{\Gamma}{\ctxwk{\Gamma}{\Gamma}}}{\idtypevar{\Gamma}}}{P}}{Q}
           \qquad
           \jterm{\ctxext{\Gamma}{\subst{\refl{\Gamma}}{{\idfunc[\Gamma]}{P}}}}{\subst{\refl{\Gamma}}{{\idfunc[\Gamma]}{Q}}}{d}}
           {\jterm{\ctxext{{{\Gamma}{\ctxwk{\Gamma}{\Gamma}}}{\idtypevar{\Gamma}}}{P}}{Q}{\tfJ(d)}}\\
& \inference
  {\jtype
    {\ctxext{{{\Gamma}{\ctxwk{\Gamma}{\Gamma}}}{\idtypevar{\Gamma}}}{P}}
    {Q}
  \qquad
  \jterm
    {\ctxext{\Gamma}{\subst{\refl{\Gamma}}{{\idfunc[\Gamma]}{P}}}}
    {\subst{\refl{\Gamma}}{{\idfunc[\Gamma]}{Q}}}
    {d}}
  {\jtermeq
    {\ctxext{{{\Gamma}{\ctxwk{\Gamma}{\Gamma}}}{\idtypevar{\Gamma}}}{P}}
    {\subst{\refl{\Gamma}}{{\idfunc[\Gamma]}{Q}}}
    {\subst{\refl{\Gamma}}{{\idfunc[\Gamma]}{\tfJ(d)}}}
    {d}}
\end{align}

@@

\begin{align}
& \inference{\jtype{\Gamma}{A}}{\jtype{\ctxext{{\Gamma}{A}}{\ctxwk{A}{A}}}{\idtypevar{A}}}\\
& \inference{\jtype{\Gamma}{A}}{\jterm{\ctxext{\Gamma}{A}}{\subst{\idfunc[A]}{\idtypevar{A}}}{\refl{A}}}\\
& \inference{\jtype{\ctxext{{{\Gamma}{A}}{\ctxwk{A}{A}}}{\idtypevar{A}}}{P}\qquad
\jterm{\ctxext{\Gamma}{A}}{\subst{\refl{A}}{{\idfunc[A]}{P}}}{d}}
{\jterm{\ctxext{{{\Gamma}{A}}{\ctxwk{A}{A}}}{\idtypevar{A}}}{P}{\tfJ(d)}}\\
& \inference{\jtype{\ctxext{{{\Gamma}{A}}{\ctxwk{A}{A}}}{\idtypevar{A}}}{P}\qquad
\jterm{\ctxext{\Gamma}{A}}{\subst{\refl{A}}{{\idfunc[A]}{P}}}{d}}
{\jtermeq
  {\ctxext{{{\Gamma}{A}}{\ctxwk{A}{A}}}{\idtypevar{A}}}
  {\subst{\refl{A}}{{\idfunc[A]}{P}}}
  {\subst{\refl{A}}{{\idfunc[A]}{\tfJ(d)}}}
  {d}}
\end{align}

@@

\subsubsection{Compatibility of identity types with extension}

\subsubsection{Compatibility of identity types with weakening}
Suppose $A$ and $B$ are types in context $\Gamma$. Then we can consider the types
\begin{align*}
\jtype*{\ctxext{{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{A}{{B}{B}}}}{\ctxwk{A}{\idtypevar{B}}}\\
\jtype*{\ctxext{{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{{A}{B}}{{A}{B}}}}{\idtypevar{\ctxwk{A}{B}}}
\end{align*}
Note that we have the judgmental equality
\begin{equation*}
\jtypeeq
  {\ctxext{{\Gamma}{A}}{\ctxwk{A}{B}}}
  {\ctxwk{A}{{B}{B}}}
  {\ctxwk{{A}{B}}{{A}{B}}}
\end{equation*}
so $\ctxwk{A}{\idtypevar{B}}$ and $\idtypevar{\ctxwk{A}{B}}$ are types in the
same context.

\begin{lem}
There is a term of type
\begin{equation*}
\jterm
  {\ctxext{{{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{{A}{B}}{{A}{B}}}}{\idtypevar{\ctxwk{A}{B}}}}
  {\ctxwk{\idtypevar{\ctxwk{A}{B}}}{\ctxwk{A}{\idtypevar{B}}}}
  {\typefont{idwktowkid}}
\end{equation*}
\end{lem}

\begin{proof}
By the identity elimination rule it suffices to find a term
\begin{equation*}
\jterm
  {\ctxext{{\Gamma}{A}}{\ctxwk{A}{B}}}
  {\subst{\refl{\ctxwk{A}{B}}}{{\idfunc[\ctxwk{A}{B}]}{\ctxwk{\idtypevar{\ctxwk{A}{B}}}{\ctxwk{A}{\idtypevar{B}}}}}}
  {i}
\end{equation*}
We may simplify the type $\subst{\refl{\ctxwk{A}{B}}}{{\idfunc[\ctxwk{A}{B}]}{\ctxwk{\idtypevar{\ctxwk{A}{B}}}{\ctxwk{A}{\idtypevar{B}}}}}$ as follows:
\begin{align*}
& \subst{\refl{\ctxwk{A}{B}}}{{\idfunc[\ctxwk{A}{B}]}{\ctxwk{\idtypevar{\ctxwk{A}{B}}}{\ctxwk{A}{\idtypevar{B}}}}}\\
& \qquad\jdeq \subst{\idfunc[\ctxwk{A}{B}]}{{\ctxwk{..}{\refl{\ctxwk{A}{B}}}}{\ctxwk{\idtypevar{\ctxwk{A}{B}}}{\ctxwk{A}{\idtypevar{B}}}}}
\end{align*}
\end{proof}

\subsubsection{Compatibility of identity types with substitution}

@@

\begin{align}
& \inference{}{\jctx{\unit}}\\
& \inference{\jctx{\Gamma}}{\jhom{\Gamma}{\unit}{\tounit{\Gamma}}}\\
& \inference{\jctx{\Gamma}\qquad\jhom{\Gamma}{\unit}{f}}{\jhomeq{\Gamma}{\unit}{f}{\tounit{\Gamma}}}
\end{align}

@@

\begin{align}
& \inference{\jctx{\Gamma}}{\jtype{\Gamma}{\unit_\Gamma}}\\
& \inference{\jctx{\Gamma}}{\jterm{\Gamma}{\unit_\Gamma}{\ttt_\Gamma}}\\
& \inference{\jtype{\ctxext{\Gamma}{\unit_\Gamma}}{P}\qquad\jterm{\Gamma}{\subst{\ttt_\Gamma}{P}}{u}}
          {\jterm{\ctxext{\Gamma}{\unit_\Gamma}}{P}{\ind{\unit_\Gamma}(u)}}\\
& \inference{\jtype{\ctxext{\Gamma}{\unit_\Gamma}}{P}\qquad\jterm{\Gamma}{\subst{\ttt_\Gamma}{P}}{u}}
          {\jtermeq{\Gamma}{\subst{\ttt_\Gamma}{P}}{\subst{\ttt_\Gamma}{\ind{\unit_\Gamma}(u)}}{u}}
\end{align}

@@

\begin{defn}
\begin{align}
& \inference{\jctx{\Gamma}}{\jctx{\tau\Gamma}}\\
& \inference{\jctx{\Gamma}}{\jhom{\Gamma}{\tau\Gamma}{t}}\\
& \inference{\jctx{\Gamma}}{\jtermeq{\ctxext{\Gamma}{\ctxwk{\Gamma}{\Gamma}}}{\ctxwk{{\Gamma}{\Gamma}}{{\Gamma}{\Delta}}}{\ctxwk{\Gamma}{t}}{\ctxwk{{\Gamma}{\Gamma}}{t}}}\\
& \inference{\jhom{\Gamma}{\Delta}{f}\qquad\jtermeq{\ctxext{\Gamma}{\ctxwk{\Gamma}{\Gamma}}}{\ctxwk{{\Gamma}{\Gamma}}{{\Gamma}{\Delta}}}{\ctxwk{\Gamma}{f}}{\ctxwk{{\Gamma}{\Gamma}}{f}}}{\jhom{\tau\Gamma}{\Delta}{\tilde{f}}}\\
& \inference{\jhom{\Gamma}{\Delta}{f}\qquad\jtermeq{\ctxext{\Gamma}{\ctxwk{\Gamma}{\Gamma}}}{\ctxwk{{\Gamma}{\Gamma}}{{\Gamma}{\Delta}}}{\ctxwk{\Gamma}{f}}{\ctxwk{{\Gamma}{\Gamma}}{f}}}
{\jtermeq{\Gamma}{\ctxwk{\Gamma}{\Delta}}{\jcomp{\Gamma}{t}{\tilde{f}}}{f}}
\end{align}
\end{defn}

@@

\subsubsection{Products}
\begin{align}
& \inference{\jctx{\Gamma}\qquad\jctx{\Delta}}{\jctx{\product{\Gamma}{\Delta}}}\\
& \inference{\jctx{\Gamma}\qquad\jctx{\Delta}}{\jhom{{\Gamma}{\Delta}}{\product{\Gamma}{\Delta}}{\pair}}
\end{align}

@@

\begin{align}
& \inference{\jctx{\Gamma}\qquad\jctx{\Delta}}{\jctx{\product{\Gamma}{\Delta}}}\\
& \inference{\jctx{\Gamma}\qquad\jctx{\Delta}}{\jhom{\product{\Gamma}{\Delta}}{\Gamma}{\proj1}}\\
& \inference{\jctx{\Gamma}\qquad\jctx{\Delta}}{\jhom{\product{\Gamma}{\Delta}}{\Delta}{\proj2}}\\
& \inference{\jhom{\greek{E}}{\Gamma}{f}\qquad\jhom{\greek{E}}{\Delta}{g}}
          {\jhom{\greek{E}}{\product{\Gamma}{\Delta}}{\pairp{f,g}}}\\
& \inference{\jhom{\greek{E}}{\Gamma}{f}\qquad\jhom{\greek{E}}{\Delta}{g}}
          {\jhomeq{\greek{E}}{\Gamma}{\jcomp{\greek{E}}{\pairp{f,g}}{\proj1}}{f}}\\
& \inference{\jhom{\greek{E}}{\Gamma}{f}\qquad\jhom{\greek{E}}{\Delta}{g}}
          {\jhomeq{\greek{E}}{\Delta}{\jcomp{\greek{E}}{\pairp{f,g}}{\proj2}}{g}}
\end{align}

\begin{equation}
\inference{\jhom{\greek{E}}{\product{\Gamma}{\Delta}}{h}
           \qquad
           \jhomeq{\greek{E}}{\Gamma}{\jcomp{E}{h}{\proj1}}{f}
           \qquad
           \jhomeq{\greek{E}}{\Delta}{\jcomp{E}{h}{\proj2}}{g}}
          {\jhomeq{\greek{E}}{\product{\Gamma}{\Delta}}{h}{\pairp{f,g}}}
\end{equation}

@@

\section{A categorical explanation of the variable-free type theory}

The categorical interpretation of type theory is a category $\catfont{C}$ of
contexts, for each object $\Gamma$ of $\catfont{C}$ a class $\catfont{type}(\Gamma)$
of morphisms into $\Gamma$ and for each $A\in\catfont{type}(\Gamma)$ a category
$\catfont{term}(A)$ of sections of $A$.
